Statistical Inconsistencies in Experimental Linguistics

### 2024-07-15 manuscript/manuscript.qmd
tided up

---
title: "Statistical Inconsistencies in Experimental Linguistics"
shorttitle: "Statistical Inconsistencies"
author:
  - name: Timo B. Roettger
    corresponding: true
    orcid: 0000-0000-0000-0001
    email: timo.roettger@gmail.com
    affiliations:
      - name: University of Oslo
        department: Department of Linguistics & Scandinavian Studies
        city: Oslo
        country: Norway
  - name: Dara Leonard Jenssen Etemady
    corresponding: false
    email: daraetemady@gmail.com
    affiliations:
      - name: University of Oslo
        department: Department of Linguistics & Scandinavian Studies
        city: Oslo
        country: Norway
author-note:
  status-changes: 
    affiliation-change: null
    deceased: null
  disclosures:
    study-registration: null
    data-sharing: null
    related-report: null
    conflict-of-interest: null
    financial-support: null
    gratitude: null
    authorship-agreements: null
abstract: "Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum."
keywords: [statistics, statcheck, reproducibility]
bibliography: bibliography.bib
format:
  apaquarto-docx: default
  apaquarto-html: default
  apaquarto-pdf:
    # Can be jou (journal), man (manuscript), stu (student), or doc (document)
    documentmode: doc
    keep-tex: true

@@

# 1. Introduction

What we know about human language and its cognitive underpinnings is often informed by experimental data, data based on which researchers draw theoretically relevant inferences using common statistical frameworks.
This inference should be transparent in order for other researchers to critically evaluate the inference process and potentially detect and correct human errors.
In the recent years, the quantitative sciences have seen repeated calls to become more transparent and reproducible by sharing data and statistical protocols [@roettger2019researcher; @arvan2022reproducibility; @laurinavichyute2022share], sharing of statistical protocols is still rare across the language sciences [@bochynska2023reproducible].
If statistical procedures cannot be critically evaluated, human errors might be left undetected and thus remain uncorrected in the publication record.
If these errors affect the decision procedure of the analysis, i.e. whether a hypothesis is rejected or accepted, these errors might lead to -at best- overconfident, -at worst- false theoretical conclusions.
The present paper will present evidence that the published literature in experimental linguistics contains a concerning amount of statistical errors, a state of affairs which warrants more rigorous data sharing practices.

>>>>

---
title: "Statistical Inconsistencies in Experimental Linguistics"
shorttitle: "Statistical Inconsistencies"
author:
  - name: Dara Leonard Jenssen Etemady
    corresponding: false
    email: daraetemady@gmail.com
    affiliations:
      - name: University of Oslo
        department: Department of Linguistics & Scandinavian Studies
        city: Oslo
        country: Norway
  - name: Timo B. Roettger
    corresponding: true
    orcid: 0000-0000-0000-0001
    email: timo.roettger@gmail.com
    affiliations:
      - name: University of Oslo
        department: Department of Linguistics & Scandinavian Studies
        city: Oslo
        country: Norway
author-note:
  status-changes: 
    affiliation-change: null
    deceased: null
  disclosures:
    study-registration: null
    data-sharing: null
    related-report: null
    conflict-of-interest: null
    financial-support: null
    gratitude: null
    authorship-agreements: null
abstract: "Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum."
keywords: [statistics, statcheck, reproducibility]
bibliography: bibliography.bib
format:
  apaquarto-docx: default
  apaquarto-html: default
  apaquarto-pdf:
    # Can be jou (journal), man (manuscript), stu (student), or doc (document)
    documentmode: doc
    keep-tex: true

@@

# 1. Introduction
What we know about human language and its cognitive underpinnings is often informed by experimental data, data based on which researchers draw theoretically relevant inferences using common statistical frameworks.
This inference should be transparent in order for other researchers to critically evaluate the inference process and potentially detect and correct human errors.
In the recent years, the quantitative sciences have seen repeated calls to become more transparent and reproducible by sharing data and statistical protocols [@roettger2019researcher; @arvan2022reproducibility; @laurinavichyute2022share], sharing of statistical protocols is still rare across the language sciences [@bochynska2023reproducible].
If statistical procedures cannot be critically evaluated, human errors might be left undetected and thus remain uncorrected in the publication record.
If these errors affect the decision procedure of the analysis, i.e. whether a hypothesis is rejected or accepted, these errors might lead to -at best- overconfident, -at worst- false theoretical conclusions.
The present paper will present evidence that the published literature in experimental linguistics contains a concerning amount of statistical errors, a state of affairs which warrants more rigorous data sharing practices.

### 2024-07-15 manuscript/manuscript.tex
tided up

\authorsnames{Timo B. Roettger,Dara Leonard Jenssen Etemady}

@@

\leftheader{Roettger and Etemady}

@@

We used the R package statcheck (Version 1.4.1-beta.2,
\citeproc{ref-nuijten2023statcheck}{Nuijten \& Epskamp, 2023}) to
automatically detect statistical-reporting inconsistencies. Statcheck
works as follows: After converting pdf or html articles to plain text,
statcheck searches for specific strings that correspond to a NHST
result, using ``regular expressions''. That way, statcheck can detect
results of t tests, F tests, Z tests, χ2 tests, correlation tests, and Q
tests as long as the test result fulfills three conditions: (a) the test
result is reported completely including the test statistic, degrees of
freedom (if applicable), and the p-value; (b) the test result is in the
body of the text, i.e.~it usually misses results reported in tables; and
(c) the test result is reported in American Psychological Association
style (APA, 2019). Given these constraints, statcheck is estimated to
detect roughly 60\% of all reported NHST results
(\citeproc{ref-nuijten2016prevalence}{Nuijten et al., 2016}) Statcheck
uses the reported test statistic and degrees of freedom to recalculate
the p-value, compares the reported and recalculated p-values and, if
there is a mismatch, reports a comparison as an ``inconsistency.'' The
algorithm takes into account that tests might have been performed as
one-tailed by identifying the search strings ``one-tailed,''
``one-sided,'' or ``directional''. Moreover, statcheck counts p = .000
and p \textless{} .000 as inconsistent as p-values of exactly zero are
mathematically impossible and the APA manual
(\citeproc{ref-APA2020}{American Psychological Association, 2020})
advises to report very small p-values as p \textless{} .001. Validity
checks suggest that inter rater reliability between manual coding and
statcheck is high, i.e.~0.76 for inconsistencies and 0.89 for decision
inconsistencies (\citeproc{ref-nuijten2016prevalence}{Nuijten et al.,
2016}). The overall accuracy of statcheck is estimated between 96.2\% to
99.9\% (\citeproc{ref-nuijten2017validity}{Nuijten et al., 2017}). We
thus consider statcheck a valid, but rough proxy of statistical
reporting inconsistencies.

>>>>

\authorsnames{Dara Leonard Jenssen Etemady,Timo B. Roettger}

@@

\leftheader{Etemady and Roettger}

@@

We used the R package statcheck (Version 1.4.1-beta.2,
\citeproc{ref-nuijten2023statcheck}{Nuijten \& Epskamp, 2023}) to
automatically detect statistical-reporting inconsistencies. Statcheck
works as follows: After converting pdf or html articles to plain text,
statcheck searches for specific strings that correspond to a NHST
result, using ``regular expressions''. That way, statcheck can detect
results of t tests, F tests, Z tests, χ2 tests, correlation tests, and Q
tests as long as the test result fulfills three conditions: (a) the test
result is reported completely including the test statistic, degrees of
freedom (if applicable), and the p-value; (b) the test result is in the
body of the text, i.e.~it usually misses results reported in tables; and
(c) the test result is reported in American Psychological Association
style (\citeproc{ref-APA2020}{American Psychological Association,
2020}). Given these constraints, statcheck is estimated to detect
roughly 60\% of all reported NHST results
(\citeproc{ref-nuijten2016prevalence}{Nuijten et al., 2016}) Statcheck
uses the reported test statistic and degrees of freedom to recalculate
the p-value, compares the reported and recalculated p-values and, if
there is a mismatch, reports a comparison as an ``inconsistency.'' The
algorithm takes into account that tests might have been performed as
one-tailed by identifying the search strings ``one-tailed,''
``one-sided,'' or ``directional''. Moreover, statcheck counts p = .000
and p \textless{} .000 as inconsistent as p-values of exactly zero are
mathematically impossible and the APA manual
(\citeproc{ref-APA2020}{American Psychological Association, 2020})
advises to report very small p-values as p \textless{} .001. Validity
checks suggest that inter rater reliability between manual coding and
statcheck is high, i.e.~0.76 for inconsistencies and 0.89 for decision
inconsistencies (\citeproc{ref-nuijten2016prevalence}{Nuijten et al.,
2016}). The overall accuracy of statcheck is estimated between 96.2\% to
99.9\% (\citeproc{ref-nuijten2017validity}{Nuijten et al., 2017}). We
thus consider statcheck a valid, but rough proxy of statistical
reporting inconsistencies.
